---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(lubridate)
```


```{r}
delayed <- clean_names(read_csv("../data/delayed-discharge-beddays-health-board.csv"))
```

Initial clean:    remove reason, remove qualifier rows


```{r}
delayed <- delayed %>% 
  select(month_of_delay,
         hbt,
         age_group,
         number_of_delayed_bed_days,
         average_daily_number_of_delayed_beds,
         reason_for_delay)
```



```{r}
delayed_no_reason <- delayed %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  select(-reason_for_delay)
```




```{r}
delayed <- delayed %>% 
  mutate(month_of_delay = ym(month_of_delay))


delayed_no_reason <- delayed_no_reason %>% 
  mutate(month_of_delay = ym(month_of_delay))
```


```{r}
unique(delayed$hbt)
```



#### Inspecting the difference over health boards (all ages)

```{r}
delayed %>% 
  filter(age_group == "18plus",
         reason_for_delay == "All Delay Reasons",
         hbt == "S08000031") %>% 
  ggplot() +
  aes(x = month_of_delay,
      y = average_daily_number_of_delayed_beds,
      group = hbt, colour = hbt) +
  geom_line(linewidth = 2, colour = "darkviolet") +
        geom_point(size = 3, colour = "darkviolet") +
      labs(title = "Average Daily Number of Delayed Beds",
           x = "\nYear",
           y = "Average Daily Number of Delayed Beds\n",
           colour = "Age Group") +
      theme(legend.position = "none",
            panel.background = element_blank(),
            panel.grid.minor.x = element_blank(),
            panel.grid.minor.y = element_blank(),
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 16),
            legend.title = element_text(size = 12),
            plot.title = element_text(size = 20))
```



Removing outlier to get a better impression of how the health boards compare

```{r}
delayed_no_reason %>% 
  slice_max(average_daily_number_of_delayed_no_reason_beds, n = 1)
```


```{r}
delayed_no_reason %>% 
  filter(age_group == "18plus",
         hbt != "S92000003") %>% 
  ggplot() +
  aes(x = month_of_delay,
      y = average_daily_number_of_delayed_beds,
      group = hbt, colour = hbt) +
  geom_line()
```

most readable graph of the year winner 2023

would be interesting to be able to look at individual health boards reactively as they seem to have
drastically different patterns to one another





#### Difference between age groups

example of difference in ages for single health board

```{r}
delayed_no_reason %>% 
  filter(hbt == "S08000015") %>% 
  ggplot() +
  aes(x = month_of_delay,
      y = average_daily_number_of_delayed_beds,
      group = age_group, colour = age_group) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
        scale_colour_manual(values = c("springgreen3", "darkblue", "darkgreen")) +
        theme(legend.position = "bottom",
            panel.background = element_blank(),
            panel.grid.minor.x = element_blank(),
            panel.grid.minor.y = element_blank(),
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 16),
            legend.title = element_text(size = 12),
            plot.title = element_text(size = 20)) +
  labs(title = "Average Daily Number of Delayed Beds",
           x = "\nYear", y = "average daily number of delayed beds\n",
       colour = "age group")
```


not super interesting, only shows the difference, maybe a line of changing proportion would be more
descriptive

```{r}
prop_change <- delayed_no_reason %>% 
  filter(age_group != "18plus") %>% 
  group_by(hbt, month_of_delay) %>%
  summarise(prop = number_of_delayed_bed_days/sum(number_of_delayed_bed_days), age_group)
```

```{r}
prop_change %>% 
  filter(age_group == "18-74",
         hbt == "S08000015") %>% 
  ggplot() +
  aes(x = month_of_delay, y = prop) +
  geom_line()
```








```{r}
delayed_no_reason %>%
  filter(hbt == "S08000015",
         age_group != "18plus") %>% 
  ggplot() +
  aes(x = month_of_delay,
      y = average_daily_number_of_delayed_beds,
      group = age_group, colour = age_group) +
  geom_line()
```




```{r}
delayed %>% 
  filter(hbt == "S08000015",
         age_group == "18plus",
         reason_for_delay != "All Delay Reasons") %>% 
  ggplot() +
  aes(x = month_of_delay, y = average_daily_number_of_delayed_beds, colour = reason_for_delay) +
  geom_line()
```



#### Hypothesis Testing


creating sets to test against one another

before = (2016-01-01   -    2020-01-01)



```{r}
test_before <- delayed_no_reason %>% 
  filter(age_group == "18plus",
         month_of_delay < "2020-01-01",
         month_of_delay >= "2017-01-01",
         hbt == "S08000022")
```

```{r}
test_after <- delayed_no_reason %>% 
  filter(age_group == "18plus",
         month_of_delay >= "2022-01-01",
         month_of_delay < "2023-01-01",
         hbt == "S08000022")
```



```{r}
t.test(test_before$number_of_delayed_bed_days,
       test_after$number_of_delayed_bed_days,
       alternative = "two.sided",
       paired = FALSE,
       conf.level = 0.95)
```





mean before and after covid

```{r}
means_before <- delayed %>% 
  filter(age_group == "18plus",
         month_of_delay < "2020-01-01",
         reason_for_delay == "All Delay Reasons"
  ) %>% 
  summarise(mean = mean(average_daily_number_of_delayed_beds), .by = hbt)
```


```{r}
means_after <- delayed %>% 
  filter(age_group == "18plus",
         month_of_delay >= "2022-01-01",
         reason_for_delay == "All Delay Reasons") %>% 
  summarise(mean = mean(average_daily_number_of_delayed_beds), .by = hbt)
```


```{r}
mean_diff <- inner_join(means_before, means_after, by = "hbt") %>% 
  mutate(mean_diff = mean.y - mean.x, HB = hbt)
```



```{r}
map <- read_csv("../data/cleaned_data/hospital_locations_clean.csv")
```

```{r}
map_means <- left_join(map, mean_diff, by = "HB")
```



```{r}
map_means %>% 
  filter(HB != "SB0802",
         HB != "SB0801") %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~ longitude,
                   lat = ~ latitude,
                   weight = 0,
                   fillColor = ~ ifelse(mean_diff > 0, "red", "green"),
                   fillOpacity = ~ abs(mean_diff)/300,
                   popup = ~ paste(Location, br(), "Board:", HB, br(), round(mean_diff, 0))
  )
```



```{r}
map_means %>% 
  filter(HB != "SB0802",
         HB != "SB0801") %>% 
  leaflet() %>% 
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addCircleMarkers(lng = ~ longitude,
                   lat = ~ latitude,
                   weight = 0,
                   fillColor = ~colorNumeric('RdYlGn', -185:185)
                   (-mean_diff),
                   fillOpacity = 0.9,
                   popup = ~ paste(Location, br(), "Board:", HB, br(), round(mean_diff, 0))
  )
```


```{r}
delayed %>%
      filter(hbt != "S92000003",
             reason_for_delay == "All Delay Reasons") %>% 
      ggplot() +
      aes(x = month_of_delay,
          y = average_daily_number_of_delayed_beds,
          group = age_group, colour = age_group) +
      geom_line() +
      geom_point(size = 1) +
      facet_wrap(~ hbt) +
      labs(title = "Average Daily Number of Delayed Beds",
           x = "\nYear",
           y = "Average Daily Number of Delayed Beds\n",
           colour = "Age Group") +
      theme(legend.position = "bottom",
            panel.background = element_blank(),
            panel.grid.minor.x = element_blank(),
            panel.grid.minor.y = element_blank(),
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 16),
            legend.title = element_text(size = 12),
            plot.title = element_text(size = 20))
```

