---
title: "Inpatient length of stay"
output: html_notebook
---

Q1: how to get to "all acute" in this data (specialties)
Q2: what's the difference in length of stay for WvS, covid?

# 1. Wrangle for all acute

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
```

```{r}
inpatients_specialty <- read_csv("../data/inpatient/inpatient_and_daycase_by_nhs_board_of_treatment_and_specialty.csv") %>% 
  clean_names() %>% 
  mutate(quarter = zoo::as.yearqtr(quarter))
```
```{r}
head(inpatients_specialty)
```

```{r}
inpatients_specialty %>% 
  group_by(specialty, specialty_name) %>% 
  summarise(count = n())
```

Which are acute?

From the [glossary](https://publichealthscotland.scot/publications/acute-hospital-activity-and-nhs-beds-information-quarterly/acute-hospital-activity-and-nhs-beds-information-quarterly-quarter-ending-30-june-2021/glossary/): 

    "'Acute' hospital care includes activity occurring in major teaching hospitals, district general hospitals and community hospitals.
    
    Includes services such as consultation with specialist clinicians; emergency treatment; routine, complex and life-saving surgery; specialist diagnostic procedures; close observation; and short-term care of patients.
    
    Excludes obstetric services; psychiatric services; long stay care services; and non-consultant led clinics."

And:

    "A specialty is defined as a division of medicine or dentistry covering a specific area of clinical activity. There are two specialty groupings (external website) that most specialties sit in: medical and surgical. The Acute specialty grouping includes the medical and surgery groupings combined."

From the 31 specialties in this dataset, even though some of these specialties deal with longer term care and/or chronic conditions (such as geriatric medicine), we are not excluding any specialties, since they are included in the broad definition of 'acute' hospital care.

# Data --> insights

## Data we have

The data we have is inpatient and day case 

* age and sex - "Age and sex breakdown for inpatient and daycase activity (episode and stays) by health board of treatment of the patients." https://www.opendata.nhs.scot/dataset/inpatient-and-daycase-activity/resource/00c00ecc-b533-426e-a433-42d79bdea5d4
* deprivation - "Deprivation breakdown for inpatient and daycase activity (episode and stays) by health board of treatment of the patients" https://www.opendata.nhs.scot/dataset/inpatient-and-daycase-activity/resource/4fc640aa-bdd4-4fbe-805b-1da1c8ed6383
* specialty - "Specialty breakdown for inpatient and daycase activity (episode and stays) by health board of treatment of the patients." https://www.opendata.nhs.scot/dataset/inpatient-and-daycase-activity/resource/c3b4be64-5fb4-4a2f-af41-b0012f0a276a

## Qs to ask

We can look at any of these to see overall activity over time, and look into winter/summer and COVID impacts:

* To what extent are the ‘winter crises’ reported by the media real?
* How has the Covid-19 pandemic affected provision of acute care in Scotland?

Looking at:

* Temporal: How has this issue changed over time? (Is it getting better or worse?)
* Geographic: How does this issue differ between areas in Scotland? (Where should efforts be focussed?)
* Demographic: Who is most affected by this issue? (Who should be targeted with efforts?)

To understand who is most impacted, we can look into age/sex and deprivation.

Some Qs from the brief:

* Are there any demographic groups that are driving the activity?
* Are there key cohorts of patient (based on their demographics) that drive the overall hospital activity?
* How significant is deprivation as a driver of hospital activity?
* Are there specialties that have more stable activity patterns over time and how much of an impact is stability on overall activity?
* How significant are the differences in hospital activity across the country? Does urban/rural have an impact or is it related to the size and capacity of the health board?

## Hospital activity by season

Look at total activity, by quarter (this is what it is grouped by)

```{r cleaning}
inpatients_age_sex <- read_csv("../data/inpatient/inpatient_and_daycase_by_nhs_board_of_treatment_age_and_sex.csv") %>% 
  clean_names()

inpatients_deprivation <- read_csv("../data/inpatient/inpatient_and_daycase_by_nhs_board_of_treatment_and_simd.csv") %>% 
  clean_names() %>% 
  # change the quarter column to yearqtr type
  mutate(quarter = zoo::as.yearqtr(quarter))
```

Interesting variables:

* admission type
* sex, age (brackets) OR deprivation quintile
* episodes, length of episode, average length of episode
* stays, length of stays, average length of stays

### admission type

```{r}
inpatients_deprivation %>% 
  distinct(admission_type)
```

### stay / spell / episode stats

Specialty data contains length of episode, length of spell
Deprivation data contains length of episode, length of stay
Sex-age data contains length of spell, length of stay

From [glossary](https://publichealthscotland.scot/publications/acute-hospital-activity-and-nhs-beds-information-quarterly/acute-hospital-activity-and-nhs-beds-information-quarterly-quarter-ending-30-june-2021/glossary/):

    "Episode
    An SMR01 episode is generated when a patient is discharged from hospital but also when a patient is transferred between hospitals, significant facilities, specialties or to the care of a different consultant.
    ...
    Average length of spell
    This is the average (mean) number of days that a patient spends in hospital during a specialty spell at a given location. It is calculated by dividing the total length of spell for all inpatients, at a given location, during a quarter by the total number of specialty spells.
    
    Average length of stay
    This is the average (mean) number of days that a patient spends in hospital during a continuous inpatient stay (CIS). It is calculated by dividing the total length of stay for all inpatients during a quarter by the total number of CISs."
    
So length of spell is time spent in one specialty. Patient may be transferred between specialties, so length of stay is total (continuous) time in hospital (in 1 or more specialty).

Not sure how length of episode fits in here (not defined in glossary).

Let's look at length of stay first. Length of spell or episode could be interesting if looking at individual specialties.

Q1 v Q3 for seasonality

```{r}
head(inpatients_deprivation)
```

```{r wrangling}
# generate stays data for all health boards to compare individual hbs to
all_hbs_stays <- inpatients_deprivation %>% 
  group_by(quarter) %>% 
  summarise(stays = sum(stays),
            length_of_stay = sum(length_of_stay),
            average_length_of_stay = length_of_stay / stays) %>% 
  mutate(hb = "All health boards", .after = quarter)

# select same columns from original df
stays_per_hb <- inpatients_deprivation %>% 
  # check group by
  group_by(hb, quarter) %>% 
  summarise(stays = sum(stays),
            length_of_stay = sum(length_of_stay),
            average_length_of_stay = length_of_stay / stays) %>% 
  select(quarter, hb, stays, length_of_stay, average_length_of_stay)

# combine all_hbs data with individual hbs data
stays <- bind_rows(all_hbs_stays, stays_per_hb)
```


```{r}
stays %>% 
  filter(hb == "All health boards") %>% 
  ggplot() +
  aes(x = quarter, y = average_length_of_stay, colour = hb) +
  geom_line() +
  geom_point()
```

try to understand what affects length of stay - look at type of admission, specialty

### drivers of length of stay (demographics, specialty)

```{r}
# average length of stay by admission type
inpatients_deprivation %>% 
  filter(average_length_of_stay < 100) %>% 
  ggplot() +
  aes(x = admission_type, y = average_length_of_stay) +
  geom_boxplot()
```

Admission type doesn't look to change average length of stay that much, just the range within each is different (not specified has longer tail, and wider IQR).

```{r}
# number of stays per SIMD by admission type over time
inpatients_deprivation %>%
  group_by(admission_type, quarter, simd) %>% 
  summarise(total_stays = sum(stays)) %>% 
  ggplot() +
  aes(x = quarter, y = total_stays, colour = admission_type) +
  geom_line() +
  facet_wrap( ~ simd, ncol = 3) +
  scale_y_continuous(labels = scales::comma_format()) +
  theme(legend.position = "bottom",
        panel.background = element_blank())
```

### episodes by specialty

```{r}
inpatients_specialty %>% 
  group_by(specialty_name) %>% 
  summarise(episodes = sum(episodes)) %>% 
  ggplot() +
  aes(y = specialty_name, x = episodes) +
  geom_col() +
  # scale_x_continuous(labels = scales::comma_format()) +
  theme(legend.position = "bottom",
        panel.background = element_blank())
```

```{r}
inpatients_specialty %>% 
  # filter(specialty_name == "General Medicine") %>%
  group_by(specialty_name, quarter) %>% 
  summarise(all_types_episodes = sum(episodes)) %>% 
  filter(all_types_episodes > 100000) %>% 
  ggplot() +
  aes(x = quarter, y = all_types_episodes, colour = specialty_name) +
  geom_line() +
  scale_y_continuous(labels = scales::comma_format()) +
  theme(legend.position = "bottom",
        panel.background = element_blank())
```

```{r}
inpatients_specialty %>% 
  filter(!specialty_name %in% c("General Medicine","General Surgery")) %>%
  group_by(specialty_name, quarter) %>% 
  summarise(all_types_episodes = sum(episodes)) %>% 
  #filter(all_types_episodes > 100000) %>% 
  ggplot() +
  aes(x = quarter, y = all_types_episodes, colour = specialty_name) +
  geom_line() +
  scale_y_continuous(labels = scales::comma_format()) +
  theme(legend.position = "bottom",
        panel.background = element_blank()) #+
  #facet_wrap(~ specialty_name, ncol = 2, scales = "free_y")
```


### who is impacted by W/S / covid patterns

Are there key patient groups driving/most affected by any seasonal pattern?

(Where do these cases come from? Look at A&E discharges "admitted to same hospital" from A&E discharge data for same time period)






