---
title: "Inpatient length of stay"
output: html_notebook
---

Q1: how to get to "all acute" in this data (specialties)
Q2: what's the difference in length of stay for WvS, covid?

# 1. Wrangle for all acute

```{r}
library(tidyverse)
library(janitor)
```

```{r}
inpatients_specialty <- read_csv("../data/inpatient/inpatient_and_daycase_by_nhs_board_of_treatment_and_specialty.csv") %>% 
  clean_names()
```
```{r}
head(inpatients_specialty)
```

```{r}
inpatients_specialty %>% 
  group_by(specialty, specialty_name) %>% 
  summarise(count = n())
```

Which are acute?

From the [glossary](https://publichealthscotland.scot/publications/acute-hospital-activity-and-nhs-beds-information-quarterly/acute-hospital-activity-and-nhs-beds-information-quarterly-quarter-ending-30-june-2021/glossary/): 

    "'Acute' hospital care includes activity occurring in major teaching hospitals, district general hospitals and community hospitals.
    
    Includes services such as consultation with specialist clinicians; emergency treatment; routine, complex and life-saving surgery; specialist diagnostic procedures; close observation; and short-term care of patients.
    
    Excludes obstetric services; psychiatric services; long stay care services; and non-consultant led clinics."

And:

    "A specialty is defined as a division of medicine or dentistry covering a specific area of clinical activity. There are two specialty groupings (external website) that most specialties sit in: medical and surgical. The Acute specialty grouping includes the medical and surgery groupings combined."

From the 31 specialties in this dataset, even though some of these specialties deal with longer term care and/or chronic conditions (such as geriatric medicine), we are not excluding any specialties, since they are included in the broad definition of 'acute' hospital care.

# Data --> insights

## Data we have

The data we have is inpatient and day case 

* age and sex - "Age and sex breakdown for inpatient and daycase activity (episode and stays) by health board of treatment of the patients." https://www.opendata.nhs.scot/dataset/inpatient-and-daycase-activity/resource/00c00ecc-b533-426e-a433-42d79bdea5d4
* deprivation - "Deprivation breakdown for inpatient and daycase activity (episode and stays) by health board of treatment of the patients" https://www.opendata.nhs.scot/dataset/inpatient-and-daycase-activity/resource/4fc640aa-bdd4-4fbe-805b-1da1c8ed6383
* specialty - "Specialty breakdown for inpatient and daycase activity (episode and stays) by health board of treatment of the patients." https://www.opendata.nhs.scot/dataset/inpatient-and-daycase-activity/resource/c3b4be64-5fb4-4a2f-af41-b0012f0a276a

## Qs to ask

We can look at any of these to see overall activity over time, and look into winter/summer and COVID impacts:

* To what extent are the ‘winter crises’ reported by the media real?
* How has the Covid-19 pandemic affected provision of acute care in Scotland?

Looking at:

* Temporal: How has this issue changed over time? (Is it getting better or worse?)
* Geographic: How does this issue differ between areas in Scotland? (Where should efforts be focussed?)
* Demographic: Who is most affected by this issue? (Who should be targeted with efforts?)

To understand who is most impacted, we can look into age/sex and deprivation.

Some Qs from the brief:

* Are there any demographic groups that are driving the activity?
* Are there key cohorts of patient (based on their demographics) that drive the overall hospital activity?
* How significant is deprivation as a driver of hospital activity?
* Are there specialties that have more stable activity patterns over time and how much of an impact is stability on overall activity?
* How significant are the differences in hospital activity across the country? Does urban/rural have an impact or is it related to the size and capacity of the health board?

## Hospital activity by season

Look at total activity, by time (quarter / month)



### Demographic impacted / drivers

Are there key patient groups driving/most affected by any seasonal pattern?

(Where do these cases come from? A&E discharges?)




