---
title: "Exploring data"
output: html_notebook
---

```{r}
library(tidyverse)
library(janitor)
library(forcats)
```

# Terminology

"chat with client"

* episode = a transfer handover event (e.g. death, discharge, move within hospital)
* acute = sharp, sudden, short-term; opposite of chronic. Acute care includes hospitals, consultations and specialisms, emergency treatment, surgeries, diagnostic procedures. Whereas non-acute = long stay service, psychiatric services

* _qf = quality of data, qualifier to note about the value in the associated variable column
* Treat SIMD as one parameter to look at, also include other demographics


# Monthly a&e data

## Discharges ----

Explore discharges data

```{r}
discharges <- read_csv("data/a_and_e/opendata_monthly_ae_discharge_202305.csv") %>% 
  clean_names()
```
```{r}
colnames(discharges)
```

```{r}
glimpse(discharges)
```

67,895 rows with 10 variables


### Looking at qf - quality notes

```{r}
discharges %>% 
  filter(!is.na(discharge_qf)) %>% 
  distinct(discharge)
```
All the _qf notes for discharge indicate a NA in discharge

```{r}
discharges %>% 
  summarise(across(.cols = everything(),
                   .fns = ~sum(is.na(.x))))
```

67,895 rows 
9,525 have NA in discharge
7,748 NA in age

Data file has 

* month
* country?
* hbt = health board code, lookup here: https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc 
* treatment_location - is this hospital?
* department_type - look at all for all A&E, but not there are departments
* age group (e.g. 18-24)
* discharge = type of discharge event (e.g. "Discharge Home...", "Transferred", "Admission...")
* number of attendances = frequency

```{r}
head(discharges)
```

### Plot discharges by age

```{r}
discharges %>% 
  mutate(age = factor(age, levels = c("Under 18", "18-24", "25-39", "40-64", "65-74", "75 plus", NA_character_))) %>% 
  ggplot() +
  aes(y = discharge, x = number_of_attendances, fill = age) +
  geom_col()
```

```{r}
min(discharges$month)
max(discharges$month)
```

Date range: 2018-01 to 2023-05

```{r}
unique(discharges$department_type)
```

## Demographics ----

```{r}
demographics <- read_csv("data/a_and_e/opendata_monthly_ae_demographics_202305.csv") %>% 
  clean_names()
```

```{r}
glimpse(demographics)
```

This gives demographics of overall attendances but we can't look at waiting times x demographics 

## Referral ----

```{r}
referral <- read_csv("data/a_and_e/opendata_monthly_ae_referral_202305.csv")
```

```{r}
glimpse(referral)
```

Referral gives how they were referred into the A&E (frequency of method)

## When ----

```{r}
when <- read_csv("data/a_and_e/opendata_monthly_ae_when_202305.csv")
```

```{r}
glimpse(when)
```

the ..._when_... data file gives the number of attendances by time of day (1h buckets, e.g. 08:00 to 08:59)


# Try mapping HBs ---

(For Ali)

```{r}
library(sf)
```

```{r}
hb_geo <- st_read(dsn = "data/map/healthboard_geo/",
                  layer = "SG_NHS_HealthBoards_2019")
```

```{r}
plot(hb_geo)
```



# Inpatient and day cases

## inpatient by age and sex

```{r}
inpatient_age_and_sex <- read_csv("data/inpatient_and_daycase_by_nhs_board_of_treatment_age_and_sex.csv")
```

```{r}
glimpse(inpatient_age_and_sex)
```

Key measures:

* Length of episode
* Length of stay
* Demographics: age, sex
* Location
* Year quarter

```{r}
# admission types
inpatient_age_and_sex %>% 
  group_by(AdmissionType) %>% 
  summarise(count = n())
```

Not sure which are acute? 

```{r}
inpatient_age_and_sex %>% 
  group_by(Quarter) %>% 
  summarise(count = n())
```

2017Q4 to 2020Q1


# Covid

## Admissions age, sex

```{r}
covid_age_sex <- read_csv("data/covid/hospital_admissions_hb_agesex_20230706.csv")
```

```{r}
glimpse(covid_age_sex)
```

* Time: week ending
* HB (but not location)
* age group, sex
* admission type and number
  * average 20182019 ascomparator
  * % variation
  
```{r}
covid_age_sex %>% 
  summarise(count = n(), .by = AdmissionType)
```

```{r}
min(covid_age_sex$WeekEnding)
max(covid_age_sex$WeekEnding)
```

5th Jan 2020 to 25 June 2023
with a comparator to same week in 2018/2019 (average of the two years)

Can look at emergency admissions and see % variation from 2018/2019 average


