---
title: "Exploring monthly A&E stats for discharges"
output: html_notebook
---

```{r}
library(tidyverse)
library(janitor)
library(forcats)
```

```{r}
discharges <- read_csv("data/a_and_e/opendata_monthly_ae_discharge_202305.csv") %>% 
  clean_names()
```
```{r}
colnames(discharges)
```

```{r}
glimpse(discharges)
```

67,895 rows with 10 variables


## Looking at qf - quality notes

```{r}
discharges %>% 
  filter(!is.na(discharge_qf)) %>% 
  distinct(discharge)
```
All the _qf notes for discharge indicate a NA in discharge

```{r}
discharges %>% 
  summarise(across(.cols = everything(),
                   .fns = ~sum(is.na(.x))))
```

67,895 rows 
9,525 have NA in discharge
7,748 NA in age

Data file has 

* month
* country?
* hbt = health board code, lookup here: https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc 
* treatment_location - is this hospital?
* department_type - look at all for all A&E, but not there are departments
* age group (e.g. 18-24)
* discharge = type of discharge event (e.g. "Discharge Home...", "Transferred", "Admission...")
* number of attendances = frequency

```{r}
head(discharges)
```

# Terminology

* episode = a transfer handover event (e.g. death, discharge, move within hospital)
* acute = sharp, sudden, short-term; opposite of chronic. Acute care includes hospitals, consultations and specialisms, emergency treatment, surgeries, diagnostic procedures. Whereas non-acute = long stay service, psychiatric services

* _qf = quality of data, qualifier to note about the value in the associated variable column
* Treat SIMD as one parameter to look at, also include other demographics


# Plot discharges by age

```{r}
discharges %>% 
  mutate(age = factor(age, levels = c("Under 18", "18-24", "25-39", "40-64", "65-74", "75 plus", NA_character_))) %>% 
  ggplot() +
  aes(y = discharge, x = number_of_attendances, fill = age) +
  geom_col()
```

```{r}
min(discharges$month)
max(discharges$month)
```

Date range: 2018-01 to 2023-05

```{r}
unique(discharges$department_type)
```


```{r}
library(sf)
```

```{r}
hb_geo <- st_read(dsn = "data/map/healthboard_geo/",
                  layer = "SG_NHS_HealthBoards_2019")
```

```{r}
plot(hb_geo)
```

