---
title: "R Notebook"
output: html_notebook
---

---
title: "R Notebook"
output: html_notebook
---
Goal: looking at waiting times at A&E
1. NumberofattendancesAll
2. Numberwithin 4 hours
3. Number over 4 hours
```{r}
waiting_times <- read.csv("../data/a_and_e/monthly_ae_activity_202305.csv")

```

```{r}
# open libraries
library(tidyverse)
library(ggplot2)
library(janitor)
library(lubridate)
```

```{r}
waiting_times_1 <- clean_names(waiting_times)
```

```{r}
#explore data
head(waiting_times)
glimpse(waiting_times)
view(waiting_times_1)
colnames(waiting_times_1)
```

```{r}
waiting_times_1 %>% 
  distinct(country)
# only a single country = Scotland

waiting_times_1 %>% 
  distinct(department_type)
# 2 departments only, injury and emergency (=A&E)


waiting_times_1 %>% 
<<<<<<< Updated upstream
  select(month, hbt, number_of_attendances_all) %>% 
  mutate(month_new = ym(month)) %>% 
  group_by(hbt) %>% 
  ggplot() +
  aes(x = month_new, group = hbt) +
  geom_histogram((aes(fill = hbt))
  
#group by hbt, and/or location, and/or department
  
  
```

=======
  select(month, hbt, number_of_attendances_all) %>%
  group_by(month) %>% 
  summarise(total_number_attendances_all = sum(number_of_attendances_all)) %>% 
  ggplot() +
  aes(x = ym(month), y = total_number_attendances_all) +
  geom_line()
  

```
```{r}
waiting_times_1 %>% 
  select(month, hbt, number_of_attendances_all) %>% 
  filter(month == str_detect(month, "^2020"))

```

```{r}
# create a line for waiting times less than 4 hours and over 4 hours

waiting_times_1 %>% 
  select(month, hbt, number_within4hours_all, number_over4hours_all) %>%
  group_by(month) %>% 
  summarise(total_number_within4hrs = sum(number_within4hours_all),
            total_number_over4hrs = sum(number_over4hours_all)) %>% 
  ggplot() +
  aes(x = ym(month)) +
  geom_line(aes(y = total_number_within4hrs), colour = "indianred") +
  geom_line(aes(y = total_number_over4hrs), colour = "springgreen4")



```
>>>>>>> Stashed changes


```{r}
#create a barplot with per month have the number of attendances in total

<<<<<<< Updated upstream
=======
```

```{r}
# trials

waiting_times_1 %>% 
  select(month, hbt, number_of_attendances_all) %>% 
  mutate(month_new = ym(month)) %>% 
  group_by(hbt) %>% 
  ggplot() +
  aes(x = month_new, group = hbt) +
  geom_line(aes(fill = hbt))
  
#group by hbt, and/or location, and/or department
>>>>>>> Stashed changes
```

